---
title: "2021 week 4 data"
author: "Megan Norman"
date: "10/18/2021"
output: github_document
---
# 2021 Bacterial Abundance Data Analysis by Megan Norman

## Load Packages

```{r load packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(lubridate)
```

## Import Data

```{r}
excel_sheets("Input_Data/week4/144L_2021_BactAbund.xlsx")

metadata <- read_excel("Input_Data/week4/144L_2021_BactAbund.xlsx", sheet = "Metadata")
glimpse(metadata)
#unique(metadata$Bottle)
#unique(metadata$Treatment)

data <- read_excel("Input_Data/week4/144L_2021_BactAbund.xlsx", sheet = "Data")
glimpse(data)

#attach data to metadata

joined <- left_join(metadata, data)
glimpse(joined)
```
## Prepare Data
We are converting the Date and Time values from characters to dates, adding columns with the time elapsed for each treatment, and converting to cells/L. Then subsetting the data for variables of interest and dropping NA values. To do this we will use the pipe operator "%>%".
```{r}
cells <- joined %>%
  mutate(Datetime = ymd_hm(Datetime),
  all_cells_L = as.numeric(all_cells_uL)*1000000,
  LNA_cells_L = as.numeric(LNA_cells_uL)*1000000,
  HNA_cells_L = as.numeric(HNA_cells_uL)*1000000) %>%
  group_by(Treatment, Bottle) %>%
  mutate(interv = interval(first(Datetime), Datetime),
         s = as.numeric(interv),
         hours = s/3600,
         days = hours/24) %>%
  ungroup() %>%
  select(Experiment:DNA_Sample, all_cells_L, LNA_cells_L, HNA_cells_L, hours, days) %>%
drop_na(all_cells_L)
glimpse(cells)
```
## Plot Growth Curves
Plotting a growth curve for each bottle with cell abundance and days data.

```{r warning=FALSE}
#assign hex colors to our different treatments.
custom.colors <- c("Control" = "#377EB8", "Kelp Exudate" = "#4DAF4A", "Kelp Exudate_Nitrate_Phosphate" = "#E41A1C", "Glucose_Nitrate_Phosphate" = "#FF7F00")

#assign levels to control order they appear in the legend
levels <- c("Control", "Kelp Exudate", "Kelp Exudate_Nitrate_Phosphate", "Glucose_Nitrate_Phosphate")

#now we will use ggplot to visualize our data.

cells %>%
  mutate(dna = ifelse(DNA_Sample == T, "*", NA)) %>%
  ggplot(aes(x=days, y=all_cells_L, group = interaction(Treatment, Bottle))) +
  geom_line(aes(color = factor(Treatment, levels = levels)), size =1) + 
  geom_point(aes(fill = factor(Treatment, levels = levels)), size = 3, color = "black", shape = 21) +
  geom_text(aes(label = dna), size = 12, color = "#E41A1C") + 
  labs(x = "Days" , y = expression(paste("Cells, L"^-1)), fill = "") +
  guides(color = "none") +
  scale_color_manual(values = custom.colors) +
  scale_fill_manual(values = custom.colors)
  #facet_grid(rows = "Treatment")
```

## Next Steps

We can calculate:
- the total change in cells from initial conditions to the end of the experiment
- specific growth rate as the slope of ln(abundance) vs time during exponential growth phase
- doubling time as ln(2) divided by the specific growth rate
- mean of each of these parameters across each treatment

First, determine *where* exponential growth is occurring in each bottle, if it does occur. To do this, plot ln(abundance) vs time.

## Identify exponential growth phase in our remin. experiments.

```{r}
ln_cells <- cells %>%
  group_by(Treatment, Bottle) %>%
  mutate(ln_cells = log(all_cells_L),
         diff_ln_cells = ln_cells - lag(ln_cells, default = first(ln_cells)))
```

Now we plot our newly calculated data.

```{r warning=FALSE}
ln_cells %>%
 mutate(dna = ifelse(DNA_Sample == T, "*", NA)) %>%
  ggplot(aes(x=days, y=diff_ln_cells, group = interaction(Treatment, Bottle))) +
  geom_line(aes(color = factor(Treatment, levels = levels)), size =1) + 
  geom_point(aes(fill = factor(Treatment, levels = levels)), size = 3, color = "black", shape = 21) +
  geom_text(aes(label = dna), size = 12, color = "#E41A1C") + 
  labs(x = "Days" , y = expression(paste("delta_ln cells, L"^-1)), fill = "") +
  guides(color = "none") +
  scale_color_manual(values = custom.colors) +
  scale_fill_manual(values = custom.colors)
```

Now, plotting ln_cells.

```{r warning=FALSE}
ln_cells %>%
 mutate(dna = ifelse(DNA_Sample == T, "*", NA)) %>%
  ggplot(aes(x=days, y=diff_ln_cells, group = interaction(Treatment, Bottle))) +
  geom_line(aes(color = factor(Treatment, levels = levels)), size =1) + 
  geom_point(aes(fill = factor(Treatment, levels = levels)), size = 3, color = "black", shape = 21) +
  geom_text(aes(label = dna), size = 12, color = "#E41A1C") + 
  labs(x = "Days" , y = expression(paste("ln cells, L"^-1)), fill = "") +
  guides(color = "none") +
  scale_color_manual(values = custom.colors) +
  scale_fill_manual(values = custom.colors) + 
  facet_wrap("Bottle", ncol =2)
```

